---
author: "Tyler Pollard, Rachel Hardy, and Hanan Ali"
date: "2024-05-06"
header-includes:
  - \usepackage{mathtools}
output:  
  github_document:
    html_preview: false
    includes: 
      in_header: _includes/head.html
    toc: true
    toc_depth: 3
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r Load Libraries}
library(knitr)
library(data.table)
library(MASS)
library(bestNormalize)
library(plyr)
library(stringr)
library(sf)
library(spData)
library(tictoc)
library(cowplot)
library(GGally)
library(patchwork)
library(fitdistrplus)
library(plotly)
library(caret)
library(splines)
library(mgcv)
library(DescTools)
library(tseries)
library(forecast)
library(car)
library(bayesplot)
library(BayesFactor)
library(rstanarm)
library(tidybayes)
library(loo)
library(brms)
library(bayesplot)
library(performance)
library(gt)
library(gtsummary)
library(tidyverse)
```

# Motivation

Coral bleaching is when corals are stressed by changes in conditions such as temperature, light, or nutrients, leading them to then expel the symbiotic algae living in their tissues, causing them to turn completely white. There are several factors that contribute to coral bleaching, including but not limited to: rising sea temperatures, rising sea levels, and ocean acidification, all of which are consequences of climate change. This study will utilize data containing covariates such as date year, latitude, longitude, temperature, turbidity, and more. The response variable for this study is the percentage of coral bleaching occurring in transect segments. The dataset contains 2,394 observations from 2006 to 2016 and is sourced from the [Florida Reef Resilience Program](https://www.bco-dmo.org/dataset/773466). \textbf{The objective of this study is to determine what covariates most heavily impact coral bleaching, and if coral bleaching has progressed over time}.

```{r Read Data}
# Clean data for all Models----
## Original Data ----
bleaching_data <- fread("_data/global_bleaching_environmental.csv", 
                        na.strings = c("", "NA", "nd"))

## Filter Data ----
## Filter to only complete Percent Bleaching and FRRP data set 
## Remove unwanted variables like temperature statistic columns
## and arrange by date for viewing purposes
filteredData <- bleaching_data |>
  filter(!is.na(Percent_Bleaching)) |>
  filter(Data_Source == "FRRP") |>
  distinct(Site_ID, Sample_ID, .keep_all = TRUE) 

## Remove unwanted variables like temperature statistic columns
## and arrange by date for viewing purposes
filteredData <- filteredData |>
  select(
    Site_ID,
    # For ordering
    Date,
    Date_Year,
    Date_Month,
    Date_Day,
    City_Town_Name,
    # Covariates
    Lat = Latitude_Degrees,
    Lon = Longitude_Degrees,
    Distance_to_Shore,
    Exposure,
    Turbidity,
    Cyclone_Frequency,
    Depth_m,
    ClimSST,
    SSTA,
    SSTA_DHW,
    TSA,
    TSA_DHW,
    Windspeed,
    # Response
    PercentBleaching = Percent_Bleaching
  ) |>
  arrange(Date)

## Remove rows with missing predictors values
bleachingData <- filteredData |>
  mutate(
    PercentBleaching = PercentBleaching/100,
    PercentBleachingBounded = case_when(
      PercentBleaching == 0 ~ 0.001,
      PercentBleaching == 1 ~ 0.999,
      .default = PercentBleaching
    ),
    PercentBleachingLow = case_when(
      PercentBleaching == 1 ~ 0.999,
      .default = PercentBleaching
    )
  )
```

# Data

The distribution for `PercentBleaching` was skewed right with support [0,1] and a relatively large number of observations (201 out of 2394) of 0 `PercentBleaching` as seen in the figure below. The beta distribution was a natural selection to model the data due to the support. After further examining the 0 values, they were deemed valid and considered as part of the same process that generated the rest of the data. To adhere to the soft (0,1) in beta regression, the 0 values were replaced with 0.001 and the 1 values were replaced by 0.999 as a simple workaround.

```{r PercentBleaching Density, fig.align='center', fig.cap="Figure 1: Density Plot of Percent Bleaching from 2,394 Coral Reef Samples"}
percentBleachingDensPlot <- ggplot(data = bleachingData) +
  geom_histogram(
    aes(x = PercentBleaching), # after_stat(density)),
    color = "#99c7c7", fill = "#bcdcdc",
    bins = 100) +
  geom_density(
    #aes(x = PercentBleaching, y = after_stat(count)),
    aes(x = PercentBleaching, y = after_stat(density)*nrow(bleachingData)/100),
    color = "#007C7C",
    linewidth = 1) +
  scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
  scale_x_continuous(breaks = seq(0,1,0.1), labels = scales::label_percent()) +
  labs(#title = "Density Plot of Percent Bleaching from 2,394 Coral\nReef Samples",
       #subtitle = "Data was Collected by the Florida Reef Resilience Program from 2006 to 2016",
       x = "Percent Bleaching",
       y = "Count") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
percentBleachingDensPlot
```

## Spatial Structure

Coral bleaching observations were **geographically distributed across Florida's reef systems**. Mapping Percent Bleaching reveals **spatial clustering**, with certain areas experiencing more severe bleaching than others. These spatial effects were captured using a **tensor product spline (`t2(Lat, Lon)`)** to model continuous variation.

```{r Spatial Structure, fig.align='center', fig.cap="Figure 2: Map of Percent Bleaching from 2,394 Coral Reef Samples around Florida"}
## Map Spatial Covariates 
county_coordinates <- map_data("county") 
spatialMap <- ggplot() +
  geom_map(
    data = county_coordinates, map = county_coordinates,
    aes(x = long, y = lat, map_id = region)
  ) +
  geom_point(
    data = bleachingData,
    aes(x = Lon, y = Lat,
        color = PercentBleaching),
    size = 0.25
  ) +
  xlim(c(-83.5,-79.5)) +
  ylim(c(24, 27.5)) +
  #facet_wrap(vars(Date_Year)) +
  scale_color_continuous(low = "green", high = "red") +
  labs(#title = "Map of Percent Bleaching from 2,394 Coral\nReef Samples around Florida",
       x = "Longitude",
       y = "Latitude") +
  theme_bw()
spatialMap
```

## Temporal Structure

The dataset spans **2006 to 2016**, allowing an analysis of **temporal bleaching trends**. Boxplots of Percent Bleaching across years show **variation over time**, with city-specific differences. To account for **nonlinear temporal trends**, we incorporated a **Gaussian Process (`gp(Date_Year, by = City_Town_Name)`)**, allowing each city to have its own temporal trend.

```{r Temporal Structure, fig.align='center', fig.cap="Figure 3: Boxplots of Percent Bleaching vs Year by City Town Name"}
## Boxplots of Percent Bleaching vs Year
yearCityPlot <- ggplot(data = bleachingData) +
  geom_boxplot(
    aes(x = factor(Date_Year), y = PercentBleaching, fill = City_Town_Name))+
    #color = "#99c7c7", fill = "#bcdcdc") +
  scale_y_continuous(breaks = seq(0,1,0.1), labels = scales::label_percent()) +
  scale_fill_discrete(name = "City Town Name", 
                      labels = c("Broward\nCounty",
                                 "Martin\nCounty",
                                 "Miami-Dade\nCounty",
                                 "Monroe\nCounty",
                                 "Palm Beach\nCounty")) +
  labs(#title = "Boxplots of Percent Bleaching vs Year by City Town Name",
       #subtitle = "Data was Collected by the Florida Reef Resilience Program from 2006 to 2016",
       x = "Date Year",
       y = "Percent Bleaching") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom"
  )
yearCityPlot
```

# Model Description

To model the percentage of coral bleaching $Y_i$, a Bayesian Beta regression with a logit link function for the mean percentage $\mu_i$ is used:

$$
\begin{aligned}
Y_i \sim Beta(\mu_i\phi, \mu_i(1-\phi))
\end{aligned}
$$

1. Model 1

2. Model 2

3. Model 3











