---
author: "Tyler Pollard, Rachel Hardy, and Hanan Ali"
date: "2024-05-06"
header-includes:
  - \usepackage{mathtools}
output:  
  github_document:
    html_preview: true
    includes: 
      in_header: _includes/head.html
    toc: true
    toc_depth: 3
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r Load Libraries}
library(knitr)
library(data.table)
library(MASS)
library(bestNormalize)
library(plyr)
library(stringr)
library(sf)
library(spData)
library(tictoc)
library(cowplot)
library(GGally)
library(patchwork)
library(fitdistrplus)
library(plotly)
library(caret)
library(splines)
library(mgcv)
library(DescTools)
library(tseries)
library(forecast)
library(car)
library(bayesplot)
library(BayesFactor)
library(rstanarm)
library(tidybayes)
library(loo)
library(brms)
library(bayesplot)
library(performance)
library(gt)
library(gtsummary)
library(tidyverse)
```

# Motivation

Coral bleaching occurs when corals experience stress due to changes in environmental conditions such as temperature, light, or nutrient levels. This stress leads corals to expel their symbiotic algae, resulting in the loss of their coloration and, in severe cases, coral death.

Several factors contribute to coral bleaching, including rising sea temperatures, sea-level changes, and ocean acidification, all of which are consequences of climate change. Understanding the key environmental drivers of bleaching is critical for conservation efforts.

The objective of this study is to identify and quantify the **impact of key environmental covariates on coral bleaching** while also assessing **how bleaching has changed over time** across different locations in Florida. Using **spatiotemporal modeling**, we analyze trends in coral bleaching by incorporating both **spatial variation** (reef locations) and **temporal patterns** (yearly changes) within a **Bayesian regression framework**.

```{r Read Data}
# Clean data for all Models----
## Original Data ----
bleaching_data <- fread("_data/global_bleaching_environmental.csv", 
                        na.strings = c("", "NA", "nd"))

## Filter Data ----
## Filter to only complete Percent Bleaching and FRRP data set 
## Remove unwanted variables like temperature statistic columns
## and arrange by date for viewing purposes
filteredData <- bleaching_data |>
  filter(!is.na(Percent_Bleaching)) |>
  filter(Data_Source == "FRRP") |>
  distinct(Site_ID, Sample_ID, .keep_all = TRUE) 

## Remove unwanted variables like temperature statistic columns
## and arrange by date for viewing purposes
filteredData <- filteredData |>
  select(
    Site_ID,
    # For ordering
    Date,
    Date_Year,
    Date_Month,
    Date_Day,
    City_Town_Name,
    # Covariates
    Lat = Latitude_Degrees,
    Lon = Longitude_Degrees,
    Distance_to_Shore,
    Exposure,
    Turbidity,
    Cyclone_Frequency,
    Depth_m,
    ClimSST,
    SSTA,
    SSTA_DHW,
    TSA,
    TSA_DHW,
    Windspeed,
    # Response
    PercentBleaching = Percent_Bleaching
  ) |>
  arrange(Date)

## Remove rows with missing predictors values
bleachingData <- filteredData |>
  mutate(
    PercentBleaching = PercentBleaching/100,
    PercentBleachingBounded = case_when(
      PercentBleaching == 0 ~ 0.001,
      PercentBleaching == 1 ~ 0.999,
      .default = PercentBleaching
    ),
    PercentBleachingLow = case_when(
      PercentBleaching == 1 ~ 0.999,
      .default = PercentBleaching
    )
  )
```

# Data

This study utilizes a dataset with 2,394 observations collected between 2006 and 2016 sourced from the [Florida Reef Resilience Program](https://www.bco-dmo.org/dataset/773466). The data includes various environmental and spatial covariates hypothesized to influence coral bleaching. The response variable, **Percent Bleaching**, measures the proportion of coral affected in each transect. Below is a list of key environmental and geographic covariates that may contribute to bleaching events: 

```{r Covariates}
covDF <- data.frame(
  Covariate = c(
    "Date_Year",
    "City_Town_Name",
    "Lat",
    "Lon",
    "Distance_to_Shore",
    "Exposure",
    "Turbidity",
    "Cyclone_Frequency",
    "Depth_m",
    "Windspeed",
    "ClimSST",
    "SSTA",
    "SSTA_DHW",
    "TSA",
    "TSA_DHW"
  ),
  Description = c(
    "Year of observation",
    "Categorical variable representing the specific city or town",
    "Latitude of the coral reef transect",
    "Longitude of the coral reef transect",
    "Distance from the reef to the shoreline (km)",
    "Level of wave exposure (e.g., sheltered, exposed)",
    "Water clarity, with higher values indicating more suspended particles",
    "Number of cyclones affecting the area per year",
    "Depth of the coral reef (meters)",
    "Average wind speed (m/s)",
    "Climatological sea surface temperature (°C)",
    "Sea surface temperature anomaly (°C)",
    "Degree heating weeks derived from SSTA",
    "Tropical sea surface temperature anomaly (°C)",
    "Degree heating weeks derived from TSA"
  )
)

covDF |>
  gt()
```

The Percent Bleaching data exhibits a right-skewed distribution (Figure 1), with a substantial number of observations reporting 0% bleaching.

```{r PercentBleaching Density, fig.align='center', fig.cap="Figure 1: Density Plot of Percent Bleaching from 2,394 Coral Reef Samples"}
percentBleachingDensPlot <- ggplot(data = bleachingData) +
  geom_histogram(
    aes(x = PercentBleaching), # after_stat(density)),
    color = "#99c7c7", fill = "#bcdcdc",
    bins = 100) +
  geom_density(
    #aes(x = PercentBleaching, y = after_stat(count)),
    aes(x = PercentBleaching, y = after_stat(density)*nrow(bleachingData)/100),
    color = "#007C7C",
    linewidth = 1) +
  scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
  scale_x_continuous(breaks = seq(0,1,0.1), labels = scales::label_percent()) +
  labs(#title = "Density Plot of Percent Bleaching from 2,394 Coral\nReef Samples",
       #subtitle = "Data was Collected by the Florida Reef Resilience Program from 2006 to 2016",
       x = "Percent Bleaching",
       y = "Count") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
percentBleachingDensPlot
```

## Spatial Structure

Coral bleaching observations were **geographically distributed across Florida's reef systems** (Figure 2). Mapping Percent Bleaching reveals **spatial clustering**, with certain areas experiencing more severe bleaching than others. 

```{r Spatial Structure, fig.align='center', fig.cap="Figure 2: Map of Percent Bleaching from 2,394 Coral Reef Samples around Florida"}
## Map Spatial Covariates 
county_coordinates <- map_data("county") 
spatialMap <- ggplot() +
  geom_map(
    data = county_coordinates, map = county_coordinates,
    aes(x = long, y = lat, map_id = region)
  ) +
  geom_point(
    data = bleachingData,
    aes(x = Lon, y = Lat,
        color = PercentBleaching),
    size = 0.25
  ) +
  xlim(c(-83.5,-79.5)) +
  ylim(c(24, 27.5)) +
  #facet_wrap(vars(Date_Year)) +
  scale_color_continuous(low = "green", high = "red") +
  labs(#title = "Map of Percent Bleaching from 2,394 Coral\nReef Samples around Florida",
       x = "Longitude",
       y = "Latitude") +
  theme_bw()
spatialMap
```

## Temporal Structure

The dataset spans **2006 to 2016**, providing an opportunity to analyze **bleaching trends over time** (Figure 3). Boxplots of Percent Bleaching over the years, categorized by City_Town_Name, reveal distinct temporal patterns across locations.

```{r Temporal Structure, fig.align='center', fig.cap="Figure 3: Boxplots of Percent Bleaching vs Year by City Town Name"}
## Boxplots of Percent Bleaching vs Year
yearCityPlot <- ggplot(data = bleachingData) +
  geom_boxplot(
    aes(x = factor(Date_Year), y = PercentBleaching, fill = City_Town_Name))+
    #color = "#99c7c7", fill = "#bcdcdc") +
  scale_y_continuous(breaks = seq(0,1,0.1), labels = scales::label_percent()) +
  scale_fill_discrete(name = "City Town Name", 
                      labels = c("Broward\nCounty",
                                 "Martin\nCounty",
                                 "Miami-Dade\nCounty",
                                 "Monroe\nCounty",
                                 "Palm Beach\nCounty")) +
  labs(#title = "Boxplots of Percent Bleaching vs Year by City Town Name",
       #subtitle = "Data was Collected by the Florida Reef Resilience Program from 2006 to 2016",
       x = "Date Year",
       y = "Percent Bleaching") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom"
  )
yearCityPlot
```

# Model Description
## Data Preprocessing

Before fitting the model, we applied several preprocessing steps:

* **Response Variable Transformation**: Since the Beta regression model requires values strictly in the (0,1) range, we replaced:

  + 0\% bleaching values with 0.001

  + 100\% bleaching values with 0.999

* **Covariate Transformations**:

  + **Yeo-Johnson transformation** was applied to all continuous covariates to reduce skewness.

  + **Centering and scaling** were performed to standardize covariates for better model convergence.
  
## Model Specification

To model the proportion of coral bleaching $Y_i$, we use a **Bayesian Beta regression** with a **logit link function**:

$$
Y_i \sim \text{Beta}(\mu_i \phi, (1-\mu_i) \phi)
$$

where $\mu_i$ is the **mean bleaching percentage**, and $\phi$ is the **precision parameter**. Various models for the mean structure were examined and defined as:

1. Model 1

$$
\text{logit}(\mu_i) = \text{Date_Year}_i\beta_1 + \text{Lat}_i\beta_2 + \text{Lon}_i\beta_3 + X_{ij}\beta_j
$$

2. Model 2

$$
\text{logit}(\mu_i) =f(\text{Date_Year}_i) + \text{Lat}_i\beta_2 + \text{Lon}_i\beta_3 + X_{ij}\beta_j
$$

3. Model 3

$$
\text{logit}(\mu_i) = f(\text{Date_Year}_i, \text{City_Town_Name}_i) + \text{Lat}_i\beta_2 + \text{Lon}_i\beta_3 + X_{ij}\beta_j
$$

4. Model 4

$$
\text{logit}(\mu_i) = \text{Date_Year}_i\beta_1 + g(\text{Lat}_i, \text{Lon}_i) + X_{ij}\beta_j
$$

5. Model 5

$$
\text{logit}(\mu_i) =f(\text{Date_Year}_i) + g(\text{Lat}_i, \text{Lon}_i) + X_{ij}\beta_j
$$

6. Model 6

$$
\text{logit}(\mu_i) = f(\text{Date_Year}_i, \text{City_Town_Name}_i) + g(\text{Lat}_i, \text{Lon}_i) + X_{ij}\beta_j
$$











