---
author: "Tyler Pollard, Rachel Hardy, and Hanan Ali"
date: "2024-05-06"
header-includes:
  - \usepackage{mathtools}
output:  
  github_document:
    html_preview: false
    #includes: 
    #  in_header: _includes/head-custom.html
    toc: true
    toc_depth: 3
    # toc_float:
    #   collapsed: false
    #   smooth_scroll: true
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r Load Libraries}
library(knitr)
library(data.table)
library(MASS)
library(bestNormalize)
library(plyr)
library(stringr)
library(sf)
library(spData)
library(tictoc)
library(cowplot)
library(GGally)
library(patchwork)
library(fitdistrplus)
library(plotly)
library(caret)
library(splines)
library(mgcv)
library(DescTools)
library(tseries)
library(forecast)
library(car)
library(bayesplot)
library(BayesFactor)
library(rstanarm)
library(tidybayes)
library(loo)
library(brms)
library(bayesplot)
library(performance)
library(gt)
library(gtsummary)
library(tidyverse)
```

# Motivation

Coral bleaching occurs when corals experience stress due to changes in environmental conditions such as temperature, light, or nutrient levels. This stress leads corals to expel their symbiotic algae, resulting in the loss of their coloration and, in severe cases, coral death.

Several factors contribute to coral bleaching, including rising sea temperatures, sea-level changes, and ocean acidification, all of which are consequences of climate change. Understanding the key environmental drivers of bleaching is critical for conservation efforts.

The objective of this study is to identify and quantify the **impact of key environmental covariates on coral bleaching** while also assessing **how bleaching has changed over time** across different locations in Florida. Using **spatiotemporal modeling**, we analyze trends in coral bleaching by incorporating both **spatial variation** (reef locations) and **temporal patterns** (yearly changes) within a **Bayesian regression framework**.

```{r Read Data}
# Clean data for all Models----
## Original Data ----
bleaching_data <- fread("_data/global_bleaching_environmental.csv", 
                        na.strings = c("", "NA", "nd"))

## Filter Data ----
## Filter to only complete Percent Bleaching and FRRP data set 
## Remove unwanted variables like temperature statistic columns
## and arrange by date for viewing purposes
filteredData <- bleaching_data |>
  filter(!is.na(Percent_Bleaching)) |>
  filter(Data_Source == "FRRP") |>
  distinct(Site_ID, Sample_ID, .keep_all = TRUE) 

## Remove unwanted variables like temperature statistic columns
## and arrange by date for viewing purposes
filteredData <- filteredData |>
  select(
    Site_ID,
    # For ordering
    Date,
    Date_Year,
    Date_Month,
    Date_Day,
    City_Town_Name,
    # Covariates
    Lat = Latitude_Degrees,
    Lon = Longitude_Degrees,
    Distance_to_Shore,
    Exposure,
    Turbidity,
    Cyclone_Frequency,
    Depth_m,
    ClimSST,
    SSTA,
    SSTA_DHW,
    TSA,
    TSA_DHW,
    Windspeed,
    # Response
    PercentBleaching = Percent_Bleaching
  ) |>
  arrange(Date)

## Remove rows with missing predictors values
bleachingData <- filteredData |>
  mutate(
    PercentBleaching = PercentBleaching/100,
    PercentBleachingBounded = case_when(
      PercentBleaching == 0 ~ 0.001,
      PercentBleaching == 1 ~ 0.999,
      .default = PercentBleaching
    ),
    PercentBleachingLow = case_when(
      PercentBleaching == 1 ~ 0.999,
      .default = PercentBleaching
    )
  )
```

# Data

This study utilizes a dataset with 2,394 observations collected between 2005 and 2016 sourced from the [Florida Reef Resilience Program](https://www.bco-dmo.org/dataset/773466). The data includes various environmental and spatial covariates hypothesized to influence coral bleaching. The response variable, **Percent Bleaching**, measures the proportion of coral affected in each transect. Below is a list of key environmental and geographic covariates that may contribute to bleaching events: 

```{r Covariates}
covDF <- data.frame(
  Covariate = c(
    "Date_Year",
    "City_Town_Name",
    "Lat",
    "Lon",
    "Distance_to_Shore",
    "Exposure",
    "Turbidity",
    "Cyclone_Frequency",
    "Depth_m",
    "Windspeed",
    "ClimSST",
    "SSTA",
    "SSTA_DHW",
    "TSA",
    "TSA_DHW"
  ),
  Description = c(
    "Year of observation",
    "Categorical variable representing the specific city or town",
    "Latitude of the coral reef transect",
    "Longitude of the coral reef transect",
    "Distance from the reef to the shoreline (km)",
    "Level of wave exposure (e.g., sheltered, exposed)",
    "Water clarity, with higher values indicating more suspended particles",
    "Number of cyclones affecting the area per year",
    "Depth of the coral reef (meters)",
    "Average wind speed (m/s)",
    "Climatological sea surface temperature (°C)",
    "Sea surface temperature anomaly (°C)",
    "Degree heating weeks derived from SSTA",
    "Tropical sea surface temperature anomaly (°C)",
    "Degree heating weeks derived from TSA"
  ),
  Units = c(
    "",
    "",
    "degrees",
    "degrees",
    "km",
    "",
    "NTU",
    "r",
    "meters",
    "m/s",
    "°C",
    "°C",
    "",
    "°C",
    ""
  )
)

covDF |>
  gt() |>
  tab_style(
    style = list(
      cell_fill(color = "#373737"),
      cell_text(color = "#fff")
    ),
    locations = list(
      cells_column_labels()
    )
  ) |>
  tab_options(
    column_labels.padding = "10px",
    table.background.color = "#f2f2f2"
  ) |>
  #opt_vertical_padding(scale = 0.25) |>
  tab_caption(caption = md("Table 1: Environmental and Geographic Covariates")) |>
  as_raw_html()
```

## Percent Bleaching Distribution

The Percent Bleaching data exhibits a right-skewed distribution (Figure 1), with a substantial number of observations reporting 0% bleaching.

```{r PercentBleaching Density, fig.align='center', fig.cap="Figure 1: Percent Bleaching Density", out.width="90%"}
percentBleachingDensPlot <- ggplot(data = bleachingData) +
  geom_histogram(
    aes(x = PercentBleaching), # after_stat(density)),
    color = "#99c7c7", fill = "#bcdcdc",
    bins = 100) +
  geom_density(
    #aes(x = PercentBleaching, y = after_stat(count)),
    aes(x = PercentBleaching, y = after_stat(density)*nrow(bleachingData)/100),
    color = "#007C7C",
    linewidth = 1) +
  scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
  scale_x_continuous(breaks = seq(0,1,0.1), labels = scales::label_percent()) +
  labs(#title = "Density Plot of Percent Bleaching from 2,394 Coral\nReef Samples",
    #subtitle = "Data was Collected by the Florida Reef Resilience Program from 2006 to 2016",
    x = "Percent Bleaching",
    y = "Count") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
percentBleachingDensPlot
```

## Spatial Structure

Coral bleaching observations were **geographically distributed across Florida's reef systems** (Figure 2). Mapping Percent Bleaching reveals **spatial clustering**, with certain areas experiencing more severe bleaching than others. 

```{r Spatial Structure, fig.align='center', fig.cap="Figure 2: Spatial Distribution of Coral Bleaching", out.width="90%"}
## Map Spatial Covariates 
county_coordinates <- map_data("county") 
spatialMap <- ggplot() +
  geom_map(
    data = county_coordinates, map = county_coordinates,
    aes(x = long, y = lat, map_id = region)
  ) +
  geom_point(
    data = bleachingData,
    aes(x = Lon, y = Lat,
        color = PercentBleaching),
    size = 0.25
  ) +
  xlim(c(-83.5,-79.5)) +
  ylim(c(24, 27.5)) +
  #facet_wrap(vars(Date_Year)) +
  #scale_color_continuous(low = "green", high = "red") +
  scale_color_gradientn(name = "Percent Bleaching",
                        colors = viridis::viridis(6)) +
  labs(#title = "Map of Percent Bleaching from 2,394 Coral\nReef Samples around Florida",
    x = "Longitude",
    y = "Latitude") +
  theme_bw() +
  theme(
    #legend.position = "bottom"
  )
spatialMap
```

## Temporal Structure

The dataset spans **2005 to 2016**, providing an opportunity to analyze **bleaching trends over time** (Figure 3). Boxplots of Percent Bleaching over the years, categorized by City_Town_Name, reveal distinct temporal patterns across locations.

```{r Temporal Structure, fig.align='center', fig.cap="Figure 3: Temporal Trends in Coral Bleaching (Boxplots)", out.width="90%"}
## Boxplots of Percent Bleaching vs Year
yearCityPlot <- ggplot(data = bleachingData) +
  geom_boxplot(
    aes(x = factor(Date_Year), y = PercentBleaching, fill = City_Town_Name))+
  #color = "#99c7c7", fill = "#bcdcdc") +
  scale_y_continuous(breaks = seq(0,1,0.1), labels = scales::label_percent()) +
  scale_fill_discrete(name = "City Town Name", 
                      labels = c("Broward\nCounty",
                                 "Martin\nCounty",
                                 "Miami-Dade\nCounty",
                                 "Monroe\nCounty",
                                 "Palm Beach\nCounty")) +
  labs(#title = "Boxplots of Percent Bleaching vs Year by City Town Name",
    #subtitle = "Data was Collected by the Florida Reef Resilience Program from 2006 to 2016",
    x = "Date Year",
    y = "Percent Bleaching") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom"
  )
yearCityPlot
```

# Model Description
## Model Specification

```{r Model Spec}
# Final Model
load(file = "_data/finalMod.RData")

iters <- 4000
burn <- 2000
chains <- 4
sims <- (iters-burn)*chains

# Get priors
finalModPriors <- get_prior(finalMod)
gpPrior <- finalModPriors |>
  filter(class == "lscale", source == "default") |>
  pull(prior) |>
  unique() |>
  str_subset(pattern = "inv_gamma")
lscalePriorAlpha <- as.numeric(str_extract(gpPrior, "(?<=inv_gamma\\()[^,]+"))
lscalePriorBeta <- as.numeric(str_extract(gpPrior, "(?<=, )[^)]+"))
```

To model the proportion of coral bleaching $Y_i$ for $i = 1, ..., 2394$, we use a **Bayesian Beta regression** with a **logit link function**:

$$
Y_i \sim \text{Beta}(\mu_i \phi, (1-\mu_i) \phi)
$$

where $\mu_i$ is the mean bleaching percentage, and $\phi$ is the precision parameter. Various models for the mean structure were examined and defined as:

<div style="padding-left:0px;padding-right:0px;padding-top:10px;padding-bottom:10px;overflow-x:auto;width:auto;">
$$
\begin{aligned}
\textbf{Model 1}: \text{logit}(\mu_i) &= \text{Date_Year}_i\beta_1 + \text{Lat}_i\beta_2 + \text{Lon}_i\beta_3 + \sum_{p} X_{ip}\beta_p \\
\textbf{Model 2}: \text{logit}(\mu_i) &= \text{Date_Year}_i\beta_1 + g(\text{Lat}, \text{Lon}) + \sum_{p} X_{ip}\beta_p \\
\textbf{Model 3}: \text{logit}(\mu_i) &= f(\text{Date_Year}) + \text{Lat}_i\beta_2 + \text{Lon}_i\beta_3 + \sum_{p} X_{ip}\beta_p \\
\textbf{Model 4}: \text{logit}(\mu_i) &= f(\text{Date_Year}) + g(\text{Lat}, \text{Lon}) + \sum_{p} X_{ip}\beta_p \\
\textbf{Model 5}: \text{logit}(\mu_i) &= f_{\text{City_Town_Name}}(\text{Date_Year}) + \text{Lat}_i\beta_2 + \text{Lon}_i\beta_3 + \sum_{p} X_{ip}\beta_p \\
\textbf{Model 6}: \text{logit}(\mu_i) &= f_{\text{City_Town_Name}}(\text{Date_Year}) + g(\text{Lat}, \text{Lon}) + \sum_{p} X_{ip}\beta_p \\
\end{aligned}
$$
</div>

where:

### Gaussian Process (GP) for Temporal Trends

$$
f_\text{City_Town_Name}(\text{Date_Year}) \sim \mathcal{GP} (0, (k_c(t_i, t_j))_{i,j = 1}^n) \\ 
$$

where the covariance function for each city $c$ is:

$$
k_c(t_i, t_j) = \sigma_c^2 \exp\left( -\frac{||t_i - t_j||^2}{2 \rho_c^2} \right)
$$

with:

* $t_i, t_j$ as observed `Date_Year` values.
* $c$ representing the city (`City_Town_Name`), where each city has a separate GP.
* $k_c(t_i, t_j)$ as the covariance function, using an exponentiated-quadratic (squared exponential) kernel.
* $\sigma_c^2$ representing a standard deviation parameter of the GP for city $c$.
* $\rho_c$ as the characteristic length-scale parameter.

### Tensor-Product Spline for Spatial Variation

$$
g(\text{Lat}, \text{Lon}) = \sum_{k_1} \sum_{k_2} \beta_{k_1 k_2} B_{k_1}(\text{Lat}) B_{k_2}(\text{Lon})
$$

where:

* $B_{k_1}(\text{Lat})$ and $B_{k_2}(\text{Lon})$ are basis functions for latitude and longitude.
* $\beta_{k_1 k_2}$ are the coefficients to be estimated.
* The smoothing penalty is controlled by a hyperparameter $\lambda$, which regularizes the estimated coefficients.

### Fixed Effects

$$
\sum_{p} X_{ip} \beta_p
$$

where:

* $X_{ip}$ is the value of the $p$-th covariate for observation $i$.
* $\beta_p$ is the corresponding regression coefficient.

### Prior Specification

* **Fixed Effects**: 
+ $\beta_p \sim \mathcal{N}(0,5)$ for all covariates $p$.
* **Gaussian Process (Temporal Trends)**:
+ $\sigma_c \sim \text{half-Cauchy}(0,2)$
+ $\rho_c \sim \text{InvGamma}(`r lscalePriorAlpha`, `r lscalePriorBeta`)$ (explicitly defined by `brms`)
* **Tensor-Product Spline**:
+ $\beta_{k_1 k_2} \sim \mathcal{N}(0,5)$
+ $\lambda \sim \text{half-Cauchy}(0,2)$ (if explicitly included in smoothing penalty)
* **Precision Parameter**: 
+ $\phi \sim \text{Gamma}(0.1, 0.1)$

This model accounts for both spatial and temporal dependencies, allowing for flexible trend estimation.

## Data Preprocessing

Before fitting the model, we applied several preprocessing steps:

* **Response Variable Transformation**: Since the Beta regression model requires values strictly in the (0,1) range, we replaced: 
+ 0\% bleaching values with 0.001
+ 100\% bleaching values with 0.999
* **Covariate Transformations**:
+ **Yeo-Johnson transformation** was applied to all continuous covariates to reduce skewness.
+ **Centering and scaling** were performed to standardize covariates for better model convergence.

# Model Comparison

We tested the 6 models abive to evaluate different approaches for capturing spatiotemporal variation in coral bleaching. The candidate models included:

- **Linear models** with Date_Year as a fixed effect.
- **GP models**, both with and without city-specific trends.
- **Smoothed Spline models**, incorporating either Lat and Lon as fixed effects or a smooth spatial term.

After running convergence checks, the final model was selected using **Leave-One-Out Cross-Validation (LOO-CV)**, ensuring it provided the best balance between fit and complexity.

```{r Model Comparison}
# Load comparison data
load(file = "_data/fitCompDF.RData")

fitCompDF |>
  select(
    Model, 
    TempStr,
    SpaceStr, 
    elpd_diff_loo, 
    se_diff_loo,
    p_loo,
    looic, 
    se_looic
  ) |>
  arrange(desc(elpd_diff_loo)) |>
  gt() |>
  cols_label(
    TempStr = "Temporal Structure",
    SpaceStr = "Spatial Structure"
    # looic = "LOOIC",
    # se_looic = "SE LOOIC",
    # p_loo = "p-value",
    # elpd_diff_loo = "LOO ELPD Difference",
    # se_diff_loo = "SE LOO Difference"
  ) |>
  fmt_number(
    use_seps = FALSE
  ) |>
  tab_footnote(
    footnote = "Difference in Expected Log pointwise Predictive Density for a new dataset",
    locations = cells_column_labels(columns = elpd_diff_loo)
  ) |>
  tab_footnote(
    footnote = "Standard Error of component-wise elpd_diff_loo between two models",
    locations = cells_column_labels(columns = se_diff_loo)
  )  |>
  tab_footnote(
    footnote = "Effective number of parameters",
    locations = cells_column_labels(columns = p_loo)
  ) |>
  tab_footnote(
    footnote = "Leave-one-out Information Criteria",
    locations = cells_column_labels(columns = looic)
  ) |>
  tab_footnote(
    footnote = "Standard Error of looic",
    locations = cells_column_labels(columns = se_looic)
  ) |>
  tab_style(
    style = cell_text(whitespace = "nowrap"),
    locations = list(cells_body(), cells_column_labels())
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#373737"),
      cell_text(color = "#fff")
    ),
    locations = list(
      cells_column_labels()
    )
  ) |>
  tab_options(
    column_labels.padding = "10px",
    table.background.color = "#f2f2f2"
  ) |>
  #opt_vertical_padding(scale = 0.25) |>
  tab_caption(caption = md("Table 2: Model Comparison  (LOO-CV)")) |>
  as_raw_html()
```

## Selected Model

Model 6 emerged as the best-performing model in the comparison based on Leave-One-Out Information Criterion (LOOIC) and expected log predictive density (ELPD). It achieved the lowest LOOIC and the highest ELPD, indicating superior predictive accuracy while effectively balancing model complexity.

A key advantage of Model 6 was its flexible structure, incorporating:

* **City-Specific GPs** for temporal variation, capturing localized trends in bleaching over time.
* A **Tensor-Product Smoothed Spline** for spatial variation, allowing for smooth, nonlinear geographic effects.
* A broad set of environmental and physical predictors, including **Distance to Shore, Exposure, Turbidity, Cyclone Frequency, Depth, Windspeed, ClimSST, SSTA, TSA, and TSA_DHW**, hypothesized to drive bleaching dynamics.

Compared to alternative models, Model 6 provided the best trade-off between fit and generalizability, avoiding overfitting while preserving essential temporal and spatial dependencies. However, some covariates exhibited credible intervals overlapping zero, suggesting they might not contribute meaningfully. To enhance interpretability and model efficiency, we performed an iterative variable selection process, systematically removing weak predictors and reassessing model performance.

## Model Refinement and Variable Selection

To improve model parsimony and predictive performance, an iterative refinement process was conducted to remove covariates that did not contribute significantly to the model. The refinement process followed these steps:

1.  Identify Non-Significant Covariates

* Variables whose 95% credible intervals contained zero were considered weak contributors.

2.  Iterative Variable Removal & Refitting

* The least significant covariate was removed from the model. 
* The model was then refit without that covariate to assess its impact. 

3.  Evaluate Model Fit via Bayes Factor & MAE

* **Bayes Factor (BF) Comparison**: The refined model was compared to the previous iteration using `bayes_factor()`. If BF > 10, the new model was preferred. 
* **LOOIC**: The reliability of how the refined model generalizes to new data was estimated. If LOOIC was lower, the new model was retained. 
* **Mean Absolute Error** (MAE): The predictive performance was evaluated using the PPD from refined model compared to observed Percent Bleaching to check model improvement/degradation. If MAE improved or remained stable, the new model was retained.

4.  Repeat Until No Further Improvement

* This process continued until all remaining covariates contributed meaningfully, ensuring the final model was both interpretable and robust.

Through this process, unnecessary covariates were systematically removed, leading to a final optimized model that retained only the most relevant predictors while maintaining strong predictive accuracy.

```{r Refinement}
load(file = "_data/refinementDF.RData")

refinementDF |>
  select(
    PriorModel, 
    RefinedModel,
    CovariateRemoved,
    BF,
    LOOIC,
    RefinedMAE
  ) |>
  gt() |>
  cols_label(
    PriorModel = "Prior Model",
    RefinedModel = "Refined Model",
    CovariateRemoved = "Covariate Removed",
    RefinedMAE = "MAE"
  ) |>
  sub_missing() |>
  fmt_number(
    columns = BF,
    decimals = 1,
    use_seps = FALSE
  ) |>
  fmt_number(
    columns = LOOIC,
    decimals = 1,
    use_seps = FALSE
  ) |>
  fmt_number(
    columns = RefinedMAE,
    decimals = 4
  ) |>
  tab_style(
    style = cell_text(whitespace = "nowrap"),
    locations = list(cells_body(), cells_column_labels())
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#373737"),
      cell_text(color = "#fff")
    ),
    locations = list(
      cells_column_labels()
    )
  ) |>
  tab_options(
    column_labels.padding = "10px",
    table.background.color = "#f2f2f2"
  ) |>
  #opt_vertical_padding(scale = 0.25) |>
  tab_caption(caption = md("Table 3: Model Refinement Results (BF, LOOIC, and MAE)")) |>
  as_raw_html()
```

## Final Model

Through iterative model comparison, Model 9 was selected as the best-performing model, showing improvements over Model 6 in terms of fit and interpretability. This selection process involved removing non-significant covariates one at a time while assessing model performance metrics. Importantly, the smoothing parameters remained unchanged throughout this refinement process, ensuring consistency in spatial and temporal trends. The final model captures essential environmental and climatic predictors, balancing complexity and generalizability.

Model 9 includes key predictors such as Distance to Shore, Turbidity, Cyclone Frequency, Windspeed, Sea Surface Temperature Anomalies (SSTA and TSA), and Degree Heating Weeks derived from TSA (TSA_DHW). These covariates were retained based on their statistical significance and their ecological relevance to coral bleaching dynamics. The refined model structure provides a robust framework for understanding and predicting bleaching patterns, facilitating targeted conservation efforts.

# Goodness of Fit

A key aspect of evaluating the selected model’s reliability is examining its ability to replicate observed data patterns. Posterior predictive checks provide a direct way to assess the extent to which simulations from the model align with the actual observed data.

## Posterior Predictive Checks

To evaluate the model’s fit, we conducted posterior predictive checks (PPCs), which compare the observed data to simulated draws from the posterior predictive distribution. The following visualizations assess whether the model-generated data resemble the observed coral bleaching percentages.

```{r PPC}
### PPC ----
# Makes draws for plots
set.seed(52) # for reproducibility
posteriorFinalMod <- posterior_predict(finalMod)

# Observed data
## Response
obsY <- bleachingData$PercentBleachingBounded

## Grouping
groupCity <- bleachingData$City_Town_Name

# Random draws
numDraws <- 400
set.seed(52) # for reproducibility
drawsInd <- sample(x = 1:sims, size = numDraws)

# Posterior sample
postYsamp <- posteriorFinalMod[drawsInd, ]

#### Plot guide 
fillPPC <- "#d1e1ec"
colorPPC <- "#b3cde0"
fill2PPC <- "#011f4b"
```

### Distribution Overlay

Figure 4 presents an overlay of the posterior predictive distribution (PPD) against the observed bleaching percentages. The solid black line represents the observed data ($y$), while the blue-shaded posterior simulations ($y_{rep}$) provide an indication of model uncertainty. The strong alignment between the observed and predicted densities suggests that the model successfully captures the overall distribution of coral bleaching percentages.

```{r PPC Distribution, fig.align='center', out.width="90%", fig.cap="Figure 4: Posterior Predictive Distribution vs Observed Data"}
#### Density ----
set.seed(52) # for reproducibility
ppcDensPlot <- ppc_dens_overlay(
  y = obsY,
  #yrep = posteriorFinalMod
  yrep = postYsamp
) +
  scale_x_continuous(
    name = "Percent Bleaching",
    limits = c(0,1),
    breaks = seq(0, 1, 0.1)
  ) +
  scale_y_continuous(
    name = "Density",
    expand = expansion(mult = c(0, 0.01))
  ) +
  labs(
    #title = "Posterior Predictive Distribution vs Observed Percent Bleaching",
    #subtitle = paste(sims, "Simulations")
  )
ppcDensPlot
```

### Distributional Statistics

The set of plots in Figure 5 evaluates how well the model reproduces key summary statistics of the observed data, including:

* Mean
* Standard deviation (SD)
* 2.5% Lower credible bound (LCB)
* Median
* 97.5% Upper credible bound (UCB)

Each histogram represents the distribution of these statistics across 8000 posterior simulations, with the vertical black line indicating the observed statistic. The Bayesian p-values assess whether the observed value is typical less than the posterior predictive distribution values. Values close to 0.5 suggest a good fit, while values near 0 or 1 may indicate potential discrepancies.

Overall, these diagnostics confirm that the final model provides a reasonable approximation of the observed data, supporting its validity for inference and prediction.

```{r PPC Stats, fig.align='center', out.width = "90%", fig.cap = "Figure 5: Posterior Predictive Checks for Summary Statistics"}
#### Stats ----
# Make stat functions
meanFunc <- function(y){mean(y)}
sdFunc <- function(y){sd(y)}
medianFunc <- function(y){median(y)}
lcbFunc <- function(y){quantile(y, 0.025)}
ucbFunc <- function(y){quantile(y, 0.975)}

##### Mean ----
set.seed(52) # for reproducibility
meanY <- meanFunc(obsY)

ppcMeanStat <- ppc_stat_data(
  y = obsY,
  yrep = posteriorFinalMod,
  group = NULL,
  stat = c("meanFunc")
) |>
  mutate(
    meanProbLow = value < meanY,
    meanProbHigh = value > meanY
  )

ppcMeanPlotGG <- ggplot() +
  geom_histogram(
    data = ppcMeanStat |> filter(variable != "y"),
    aes(x = value, color = "Posterior"),
    fill = fillPPC
  ) +
  geom_vline(
    data = ppcMeanStat |> filter(variable == "y"),
    aes(xintercept = value, color = "Observed"),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = "Percent Bleaching"
  ) +
  scale_y_continuous(
    name = "Number of Posterior Draws",
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_color_manual(
    name = "Data",
    values = c(
      "Posterior" = colorPPC,
      "Observed" = "black"
    ),
    breaks = c("Posterior", "Observed")
  ) +
  labs(title = "Mean",
       subtitle = paste("p-value =", round(mean(ppcMeanStat$meanProbLow[-1]), 4))
  ) +
  theme_bw() +
  guides(color = guide_legend(byrow = TRUE)) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.key.spacing.y = unit(5, "points"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

##### SD ----
set.seed(52) # for reproducibility
sdY <- sdFunc(obsY)

ppcSDStat <- ppc_stat_data(
  y = obsY,
  yrep = posteriorFinalMod,
  group = NULL,
  stat = c("sdFunc")
) |>
  mutate(
    sdProbLow = value < sdY,
    sdProbHigh = value > sdY
  )

ppcSDPlotGG <- ggplot() +
  geom_histogram(
    data = ppcSDStat |> filter(variable != "y"),
    aes(x = value, color = "Posterior"),
    fill = fillPPC
  ) +
  geom_vline(
    data = ppcSDStat |> filter(variable == "y"),
    aes(xintercept = value, color = "Observed"),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = "Percent Bleaching"
  ) +
  scale_y_continuous(
    name = "Number of Posterior Draws",
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_color_manual(
    name = "Data",
    values = c(
      "Posterior" = colorPPC,
      "Observed" = "black"
    ),
    breaks = c("Posterior", "Observed")
  ) +
  labs(title = "SD",
       subtitle = paste("p-value =", round(mean(ppcSDStat$sdProbLow[-1]), 4))
  ) +
  theme_bw() +
  guides(color = guide_legend(byrow = TRUE)) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.key.spacing.y = unit(5, "points"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

##### Median ----
set.seed(52) # for reproducibility
medianY <- medianFunc(obsY)

ppcMedianStat <- ppc_stat_data(
  y = obsY,
  yrep = posteriorFinalMod,
  group = NULL,
  stat = c("medianFunc")
) |>
  mutate(
    medianProbLow = value < medianY,
    medianProbHigh = value > medianY
  )

ppcMedianPlotGG <- ggplot() +
  geom_histogram(
    data = ppcMedianStat |> filter(variable != "y"),
    aes(x = value, color = "Posterior"),
    fill = fillPPC
  ) +
  geom_vline(
    data = ppcMedianStat |> filter(variable == "y"),
    aes(xintercept = value, color = "Observed"),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = "Percent Bleaching"
  ) +
  scale_y_continuous(
    name = "Number of Posterior Draws",
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_color_manual(
    name = "Data",
    values = c(
      "Posterior" = colorPPC,
      "Observed" = "black"
    ),
    breaks = c("Posterior", "Observed")
  ) +
  labs(title = "Median",
       subtitle = paste("p-value =", round(mean(ppcMedianStat$medianProbLow[-1]), 4))
  ) +
  theme_bw() +
  guides(color = guide_legend(byrow = TRUE)) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.key.spacing.y = unit(5, "points"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

##### LCB ----
set.seed(52) # for reproducibility
lcbY <- lcbFunc(obsY)

ppcLCBStat <- ppc_stat_data(
  y = obsY,
  yrep = posteriorFinalMod,
  group = NULL,
  stat = c("lcbFunc")
) |>
  mutate(
    lcbProbLow = value < lcbY,
    lcbProbHigh = value > lcbY
  )

ppcLCBPlotGG <- ggplot() +
  geom_histogram(
    data = ppcLCBStat |> filter(variable != "y"),
    aes(x = value, color = "Posterior"),
    fill = fillPPC
  ) +
  geom_vline(
    data = ppcLCBStat |> filter(variable == "y"),
    aes(xintercept = value, color = "Observed"),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = "Percent Bleaching"
  ) +
  scale_y_continuous(
    name = "Number of Posterior Draws",
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_color_manual(
    name = "Data",
    values = c(
      "Posterior" = colorPPC,
      "Observed" = "black"
    ),
    breaks = c("Posterior", "Observed")
  ) +
  labs(title = "LCB",
       subtitle = paste("p-value =", round(mean(ppcLCBStat$lcbProbLow[-1]), 4))
  ) +
  theme_bw() +
  guides(color = guide_legend(byrow = TRUE)) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.key.spacing.y = unit(5, "points"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

##### UCB ----
set.seed(52) # for reproducibility
ucbY <- ucbFunc(obsY)

ppcUCBStat <- ppc_stat_data(
  y = obsY,
  yrep = posteriorFinalMod,
  group = NULL,
  stat = c("ucbFunc")
) |>
  mutate(
    ucbProbLow = value < ucbY,
    ucbProbHigh = value > ucbY
  )

ppcUCBPlotGG <- ggplot() +
  geom_histogram(
    data = ppcUCBStat |> filter(variable != "y"),
    aes(x = value, color = "Posterior"),
    fill = fillPPC
  ) +
  geom_vline(
    data = ppcUCBStat |> filter(variable == "y"),
    aes(xintercept = value, color = "Observed"),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = "Percent Bleaching"
  ) +
  scale_y_continuous(
    name = "Number of Posterior Draws",
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_color_manual(
    name = "Data",
    values = c(
      "Posterior" = colorPPC,
      "Observed" = "black"
    ),
    breaks = c("Posterior", "Observed")
  ) +
  labs(title = "UCB",
       subtitle = paste("p-value =", round(mean(ppcUCBStat$ucbProbLow[-1]), 4))
  ) +
  theme_bw() +
  guides(color = guide_legend(byrow = TRUE)) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.key.spacing.y = unit(5, "points"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

#### Combined plot ----
ppcCombPlot <- 
  (ppcMeanPlotGG + ppcSDPlotGG) /
  (ppcLCBPlotGG + ppcMedianPlotGG + ppcUCBPlotGG) +
  plot_layout(
    guides = "collect",
    axes = "collect_x"
  ) +
  plot_annotation(
    #title = "Posterior Predictive Checks for Distributional Statistics",
    #subtitle = paste("Bayesian predictive p-values for", sims, "Simulations"),
    theme = theme(
      legend.position = "bottom",
      legend.direction = "horizontal"
    )
  )
ppcCombPlot
```

# Model Results

After validating the model's performance through posterior predictive checks, we now examine the key results. This section explores the significance of model predictors, the temporal trends in bleaching events, and the spatial distribution of bleaching risk across Florida's coral reefs.

## Variable Importance

The table below presents the estimated fixed effects from the final Bayesian Beta regression model. Each coefficient represents the effect of a predictor on the proportion of coral bleaching. The interpretation of key predictors is as follows:

```{r Model Results Table}
fixedEff <- fixef(finalMod) |>
  data.frame() |>
  mutate(
    p_val = dnorm(Estimate/Est.Error)
  ) |>
  mutate(
    across(everything(), function(x){round(x, 4)})
  ) |>
  mutate(
    Sig = ifelse(p_val < 0.01, "***",
                 ifelse(p_val < 0.05, "**",
                        ifelse(p_val < 0.1, "*", "")))
  )

# Render Table
fixedEff |>
  rownames_to_column(var = "Parameter") |>
  select(
    Parameter,
    Estimate,
    Est.Error,
    Q2.5,
    Q97.5
  ) |>
  filter(str_detect(Parameter, "t2", negate = TRUE)) |>
  gt() |>
  cols_label(
    Estimate = "{{:beta:}}",
    Est.Error = "SD({{:beta:}})"
  ) |>
  cols_merge(
    columns = c(Q2.5, Q97.5), pattern = "({1}, {2})"
  ) |>
  cols_label(
    Q2.5 = "95% CI"
  ) |>
  cols_align(
    columns = Q2.5,
    align = "center"
  ) |>
  tab_footnote(
    footnote = "Parameter estimate",
    locations = cells_column_labels(columns = Estimate)
  ) |>
  tab_footnote(
    footnote = "Standard Deviation of parameter estimate",
    locations = cells_column_labels(columns = Est.Error)
  ) |>
  tab_footnote(
    footnote = "95% Credible Interval of parameter estimate",
    locations = cells_column_labels(columns = Q2.5)
  ) |>
  tab_style(
    style = cell_text(whitespace = "nowrap"),
    locations = list(cells_body(), cells_column_labels())
  ) |>
  tab_style(
    style = list(
      cell_text(
        align = "center"
      )
    ),
    locations = list(
      cells_column_labels(
        columns = c(Estimate, Est.Error)
      )
    )
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#373737"),
      cell_text(color = "#fff")
    ),
    locations = list(
      cells_column_labels()
    )
  ) |>
  tab_options(
    column_labels.padding = "10px",
    table.background.color = "#f2f2f2"
  ) |>
  #opt_vertical_padding(scale = 0.25) |>
  tab_caption(caption = md("Table 4: Estimated Fixed Effects from Bayesian Beta Regression")) |> 
  as_raw_html()
```

### Key Observations:

* Distance to Shore (β = 0.1005, 95% CI: [0.0338, 0.1660]) – Positively associated with bleaching, indicating that reefs farther from shore experience slightly higher bleaching, potentially due to differences in water quality and exposure to open ocean stressors.

* Turbidity (β = -0.0784, 95% CI: [-0.1270, -0.0295]) – Negatively associated with bleaching, suggesting that murkier waters may provide some shielding from temperature-induced stress.

* Cyclone Frequency & Wind Speed (β = -0.0524, -0.0466) – Moderate negative effects, likely due to increased mixing of ocean layers, reducing localized heat stress on corals.

* Sea Surface Temperature Anomaly (SSTA) (β = -0.0573, 95% CI: [-0.1104, -0.0047]) – Contrary to expectations, this predictor has a small negative effect, possibly reflecting interactions with other environmental conditions or non-linear temperature effects.

* Thermal Stress Accumulation (TSA) & Degree Heating Weeks (TSA_DHW) (β = 0.1313, 0.0885) – Significant positive effects, confirming that prolonged heat stress increases bleaching probability.

## Temporal Effects

To evaluate how bleaching trends evolve over time, we analyze posterior estimates of the temporal effect from the GP component.

### County-Specific Trends

The plot below illustrates the estimated temporal variation in bleaching probability across five Florida counties from 2006 to 2016:

```{r Temporal Effects Facet, fig.align='center', out.width="90%", fig.cap="Figure 6: Temporal Effects by County (Smoothed Trends)"}

load(file = "_data/condEffDataTemporal.RData")
condEffDataTemporal <- condEffDataTemporal$Date_Year

##### Facet ----
condEffDataTemporalGGfacet <-
  ggplot() +
  geom_point(
    data = bleachingData,
    aes(x = Date_Year, y = PercentBleachingBounded),
    #color = City_Town_Name)
    alpha = 0.2
  ) +
  geom_ribbon(
    data = condEffDataTemporal2,
    aes(x = Date_Year, ymin = lower__, ymax = upper__,
        fill = City_Town_Name), 
    alpha = 0.3
  ) +
  geom_line(
    data = condEffDataTemporal2,
    aes(x = Date_Year, y = estimate__, 
        color = City_Town_Name),
    linewidth = 1.5
  ) +
  facet_wrap(vars(City_Town_Name)) +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.1), 
                     labels = scales::label_percent()) +
  scale_x_continuous(limits = range(bleachingData$Date_Year),
                     breaks = seq(min(bleachingData$Date_Year),
                                  max(bleachingData$Date_Year),
                                  1)
  ) +
  labs(
    x = "Date Year",
    y = "Percent Bleaching"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor.x = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5),
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.2),
    legend.justification = c(0.25, 0.5)
  )

condEffDataTemporalGGfacet
```

#### Key Findings:

* Each county shows unique bleaching patterns, with different peak years.

* Monroe and Miami-Dade counties have higher bleaching probabilities and greater interannual variability than the others.

* Martin County shows a steady increase after 2010, suggesting it may be particularly vulnerable to long-term climate stress.

This faceted plot illustrates the modeled county-specific temporal variation in bleaching probability, capturing how bleaching risk fluctuates over time in different locations.

### Overlaid Trends

To provide a broader comparison of modeled bleaching trends across counties, the following plot presents an overlay of the estimated temporal effects without faceting:

```{r Temporal Effects Overlay, fig.align='center', out.width="90%", fig.cap="Figure 7: Overlaid Temporal Trends in Percent Bleaching"}

##### Overlay ----
condEffDataTemporalGGoverlay <-
  ggplot() +
  # geom_point(
  #   data = bleachingData,
  #   aes(x = Date_Year, y = PercentBleachingBounded, 
  #       color = City_Town_Name)
  # ) +
  geom_line(
    data = condEffDataTemporal2,
    aes(x = Date_Year, y = estimate__, 
        color = City_Town_Name),
    linewidth = 1
  ) +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.1), 
                     labels = scales::label_percent()) +
  scale_x_continuous(breaks = unique(bleachingData$Date_Year)) +
  labs(
    x = "Date Year",
    y = "Percent Bleaching"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor.x = element_blank()
  )

condEffDataTemporalGGoverlay
```

#### Key Observations:

* The overlay allows for direct comparison between counties, highlighting relative differences in bleaching probabilities.

* Monroe County stands out with the most extreme bleaching peak in 2014-2015.

* Palm Beach County maintains relatively low bleaching levels compared to other regions.

* The synchronized bleaching peaks across counties suggest that widespread regional environmental drivers, such as temperature anomalies, are at play.

While this plot removes individual county facets, it retains the key modeled trends and provides a clearer comparative perspective on bleaching severity across regions.

## Spatial Effects

The spatial effects plot provides insights into regional differences in bleaching susceptibility. The spatial random effect was modeled using a tensor-product spline, capturing localized variations beyond the fixed effects.

```{r Spatial Effects, fig.align='center', out.width="90%", fig.cap="Figure 8: Spatial Effects of Coral Bleaching (Modeled Estimates)"}

load(file = "_data/condSmoothData.RData")
condSmoothData <- condSmoothData$`mu: t2(Lon,Lat`
condLonRange <- range(condSmoothData2$Lon)
condLatRange <- range(condSmoothData2$Lat)

county_coordinates <- map_data("county") 
condSmoothPlotSpatialGG <- ggplot() +
  geom_raster(
    data = condSmoothData2,
    aes(x = Lon, y = Lat, fill = estimate__)
  ) +
  # geom_point(
  #   data = bleachingData,
  #   aes(x = Lon, y = Lat,
  #       color = PercentBleaching),
  #   size = 0.25, alpha = 0.1
  # ) +
  geom_map(
    data = county_coordinates, map = county_coordinates,
    aes(x = long, y = lat, map_id = region)
  ) +
  scale_fill_gradientn(name = "Spatial Estimate",
                       colors = viridis::viridis(6)) +
  # scale_color_gradientn(name = "Spatial Estimate",
  #                       colors = viridis::viridis(6)) +
  #scale_color_continuous(low = "green", high = "red") +
  xlim(condLonRange) +
  ylim(condLatRange) +
  theme_bw()
condSmoothPlotSpatialGG 
```

### Key Findings:

* Higher bleaching probabilities are concentrated in nearshore reefs along the Florida Keys and southeastern coastline, reinforcing the importance of local environmental stressors.

* Some offshore reef systems exhibit lower bleaching risk, potentially due to deeper waters or localized upwelling that buffers temperature stress.

* The spatial gradient suggests that conservation efforts should prioritize areas with high predicted bleaching risk, particularly in the southeastern coastal zone.

# Discussion

This study applied a spatiotemporal modeling approach to assess coral bleaching trends across Florida’s reef systems, capturing both geographic variation and temporal changes. The results demonstrate that prolonged temperature anomalies (**TSA** and **TSA_DHW**) are the strongest predictors of bleaching, with additional associations observed for turbidity, cyclone frequency, and wind speed. While thermal stress is well-documented as a primary driver of bleaching, this analysis suggests that local environmental conditions, such as water quality and storm activity, may influence bleaching severity in complex ways.

By incorporating both spatial and temporal variation, the model identifies region-specific and time-dependent patterns of bleaching risk, reinforcing the importance of localized conservation strategies. These findings emphasize the need for continued monitoring and adaptive management that considers both large-scale climate stressors and site-specific environmental factors.

## Limitations

While this study provides valuable insights, several limitations should be considered:

* **Data Constraints**: The dataset spans from 2006 to 2016, preventing assessment of more recent bleaching trends.

* **Model Assumptions**: The Bayesian Beta regression model imposes distributional constraints that may not fully capture extreme bleaching events.

* **Unmeasured Factors**: Variables such as nutrient levels, local anthropogenic impacts, and additional reef health indicators were not included but may play a role in bleaching dynamics.

## Future Directions

While this study leveraged the most recent available FRRP data (2005–2016), future research could benefit from new data collection to assess whether the observed trends persist under current climate conditions. Additionally, further exploration of existing datasets could provide deeper insights into bleaching patterns by incorporating complementary environmental variables or alternative modeling approaches.

Potential areas for methodological refinement include:

* **Investigating Local Influences**: Further analysis could assess whether turbidity consistently mitigates bleaching severity and how storm-induced ocean mixing interacts with thermal stress.

* **Enhancing Modeling Approaches**: Exploring alternative statistical frameworks, such as hierarchical or machine learning-based models, may improve predictive accuracy and better capture nonlinear relationships.

* **Integrating Additional Environmental Variables**: If future datasets allow, incorporating nutrient levels, pollution metrics, or additional reef health indicators could refine understanding of bleaching dynamics.

By focusing on these methodological improvements, future research can build upon this study’s findings to further improve bleaching risk assessments and conservation planning.
