---
author: "Tyler Pollard, Rachel Hardy, and Hanan Ali"
date: "2024-05-06"
header-includes:
  - \usepackage{mathtools}
output:  
  github_document:
    html_preview: false
    includes: 
      in_header: _includes/head.html
    toc: true
    toc_depth: 3
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r Load Libraries}
library(knitr)
library(data.table)
library(MASS)
library(bestNormalize)
library(plyr)
library(stringr)
library(sf)
library(spData)
library(tictoc)
library(cowplot)
library(GGally)
library(patchwork)
library(fitdistrplus)
library(plotly)
library(caret)
library(splines)
library(mgcv)
library(DescTools)
library(tseries)
library(forecast)
library(car)
library(bayesplot)
library(BayesFactor)
library(rstanarm)
library(tidybayes)
library(loo)
library(brms)
library(bayesplot)
library(performance)
library(gt)
library(gtsummary)
library(tidyverse)
```

# Motivation

Coral bleaching occurs when corals experience stress due to changes in environmental conditions such as temperature, light, or nutrient levels. This stress leads corals to expel their symbiotic algae, resulting in the loss of their coloration and, in severe cases, coral death.

Several factors contribute to coral bleaching, including rising sea temperatures, sea-level changes, and ocean acidification, all of which are consequences of climate change. Understanding the key environmental drivers of bleaching is critical for conservation efforts.

The objective of this study is to identify and quantify the **impact of key environmental covariates on coral bleaching** while also assessing **how bleaching has changed over time** across different locations in Florida. Using **spatiotemporal modeling**, we analyze trends in coral bleaching by incorporating both **spatial variation** (reef locations) and **temporal patterns** (yearly changes) within a **Bayesian regression framework**.

```{r Read Data}
# Clean data for all Models----
## Original Data ----
bleaching_data <- fread("_data/global_bleaching_environmental.csv", 
                        na.strings = c("", "NA", "nd"))

## Filter Data ----
## Filter to only complete Percent Bleaching and FRRP data set 
## Remove unwanted variables like temperature statistic columns
## and arrange by date for viewing purposes
filteredData <- bleaching_data |>
  filter(!is.na(Percent_Bleaching)) |>
  filter(Data_Source == "FRRP") |>
  distinct(Site_ID, Sample_ID, .keep_all = TRUE) 

## Remove unwanted variables like temperature statistic columns
## and arrange by date for viewing purposes
filteredData <- filteredData |>
  select(
    Site_ID,
    # For ordering
    Date,
    Date_Year,
    Date_Month,
    Date_Day,
    City_Town_Name,
    # Covariates
    Lat = Latitude_Degrees,
    Lon = Longitude_Degrees,
    Distance_to_Shore,
    Exposure,
    Turbidity,
    Cyclone_Frequency,
    Depth_m,
    ClimSST,
    SSTA,
    SSTA_DHW,
    TSA,
    TSA_DHW,
    Windspeed,
    # Response
    PercentBleaching = Percent_Bleaching
  ) |>
  arrange(Date)

## Remove rows with missing predictors values
bleachingData <- filteredData |>
  mutate(
    PercentBleaching = PercentBleaching/100,
    PercentBleachingBounded = case_when(
      PercentBleaching == 0 ~ 0.001,
      PercentBleaching == 1 ~ 0.999,
      .default = PercentBleaching
    ),
    PercentBleachingLow = case_when(
      PercentBleaching == 1 ~ 0.999,
      .default = PercentBleaching
    )
  )
```

# Data

This study utilizes a dataset with 2,394 observations collected between 2006 and 2016 sourced from the [Florida Reef Resilience Program](https://www.bco-dmo.org/dataset/773466). The data includes various environmental and spatial covariates hypothesized to influence coral bleaching. The response variable, **Percent Bleaching**, measures the proportion of coral affected in each transect. Below is a list of key environmental and geographic covariates that may contribute to bleaching events: 

```{r Covariates}
covDF <- data.frame(
  Covariate = c(
    "Date_Year",
    "City_Town_Name",
    "Lat",
    "Lon",
    "Distance_to_Shore",
    "Exposure",
    "Turbidity",
    "Cyclone_Frequency",
    "Depth_m",
    "Windspeed",
    "ClimSST",
    "SSTA",
    "SSTA_DHW",
    "TSA",
    "TSA_DHW"
  ),
  Description = c(
    "Year of observation",
    "Categorical variable representing the specific city or town",
    "Latitude of the coral reef transect",
    "Longitude of the coral reef transect",
    "Distance from the reef to the shoreline (km)",
    "Level of wave exposure (e.g., sheltered, exposed)",
    "Water clarity, with higher values indicating more suspended particles",
    "Number of cyclones affecting the area per year",
    "Depth of the coral reef (meters)",
    "Average wind speed (m/s)",
    "Climatological sea surface temperature (°C)",
    "Sea surface temperature anomaly (°C)",
    "Degree heating weeks derived from SSTA",
    "Tropical sea surface temperature anomaly (°C)",
    "Degree heating weeks derived from TSA"
  )
)

covDF |>
  gt() |>
  tab_style(
    style = list(
      cell_fill(color = "#373737"),
      cell_text(color = "#fff")
    ),
    locations = list(
      cells_column_labels()
    )
  ) |>
  tab_options(
    column_labels.padding = "10px",
    table.background.color = "#f2f2f2"
  ) |>
  #opt_vertical_padding(scale = 0.25) |>
  tab_caption(caption = md("Table 1: Covariate Descriptions")) |>
  as_raw_html()
```

The Percent Bleaching data exhibits a right-skewed distribution (Figure 1), with a substantial number of observations reporting 0% bleaching.

```{r PercentBleaching Density, fig.align='center', fig.cap="Figure 1: Density Plot of Percent Bleaching from 2,394 Coral Reef Samples", out.width="90%"}
percentBleachingDensPlot <- ggplot(data = bleachingData) +
  geom_histogram(
    aes(x = PercentBleaching), # after_stat(density)),
    color = "#99c7c7", fill = "#bcdcdc",
    bins = 100) +
  geom_density(
    #aes(x = PercentBleaching, y = after_stat(count)),
    aes(x = PercentBleaching, y = after_stat(density)*nrow(bleachingData)/100),
    color = "#007C7C",
    linewidth = 1) +
  scale_y_continuous(expand = expansion(mult = c(0,0.05))) +
  scale_x_continuous(breaks = seq(0,1,0.1), labels = scales::label_percent()) +
  labs(#title = "Density Plot of Percent Bleaching from 2,394 Coral\nReef Samples",
    #subtitle = "Data was Collected by the Florida Reef Resilience Program from 2006 to 2016",
    x = "Percent Bleaching",
    y = "Count") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
percentBleachingDensPlot
```

## Spatial Structure

Coral bleaching observations were **geographically distributed across Florida's reef systems** (Figure 2). Mapping Percent Bleaching reveals **spatial clustering**, with certain areas experiencing more severe bleaching than others. 

```{r Spatial Structure, fig.align='center', fig.cap="Figure 2: Map of Percent Bleaching from 2,394 Coral Reef Samples around Florida", out.width="90%"}
## Map Spatial Covariates 
county_coordinates <- map_data("county") 
spatialMap <- ggplot() +
  geom_map(
    data = county_coordinates, map = county_coordinates,
    aes(x = long, y = lat, map_id = region)
  ) +
  geom_point(
    data = bleachingData,
    aes(x = Lon, y = Lat,
        color = PercentBleaching),
    size = 0.25
  ) +
  xlim(c(-83.5,-79.5)) +
  ylim(c(24, 27.5)) +
  #facet_wrap(vars(Date_Year)) +
  scale_color_continuous(low = "green", high = "red") +
  labs(#title = "Map of Percent Bleaching from 2,394 Coral\nReef Samples around Florida",
    x = "Longitude",
    y = "Latitude") +
  theme_bw() +
  theme(
    #legend.position = "bottom"
  )
spatialMap
```

## Temporal Structure

The dataset spans **2006 to 2016**, providing an opportunity to analyze **bleaching trends over time** (Figure 3). Boxplots of Percent Bleaching over the years, categorized by City_Town_Name, reveal distinct temporal patterns across locations.

```{r Temporal Structure, fig.align='center', fig.cap="Figure 3: Boxplots of Percent Bleaching vs Year by City Town Name", out.width="90%"}
## Boxplots of Percent Bleaching vs Year
yearCityPlot <- ggplot(data = bleachingData) +
  geom_boxplot(
    aes(x = factor(Date_Year), y = PercentBleaching, fill = City_Town_Name))+
  #color = "#99c7c7", fill = "#bcdcdc") +
  scale_y_continuous(breaks = seq(0,1,0.1), labels = scales::label_percent()) +
  scale_fill_discrete(name = "City Town Name", 
                      labels = c("Broward\nCounty",
                                 "Martin\nCounty",
                                 "Miami-Dade\nCounty",
                                 "Monroe\nCounty",
                                 "Palm Beach\nCounty")) +
  labs(#title = "Boxplots of Percent Bleaching vs Year by City Town Name",
    #subtitle = "Data was Collected by the Florida Reef Resilience Program from 2006 to 2016",
    x = "Date Year",
    y = "Percent Bleaching") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom"
  )
yearCityPlot
```

# Model Description
## Data Preprocessing

Before fitting the model, we applied several preprocessing steps:

* **Response Variable Transformation**: Since the Beta regression model requires values strictly in the (0,1) range, we replaced: 
  + 0\% bleaching values with 0.001
  + 100\% bleaching values with 0.999
* **Covariate Transformations**:
  + **Yeo-Johnson transformation** was applied to all continuous covariates to reduce skewness.
  + **Centering and scaling** were performed to standardize covariates for better model convergence.

## Model Specification

```{r Model Spec}
# Final Model
load(file = "_data/finalMod.RData")

iters <- 4000
burn <- 2000
chains <- 4
sims <- (iters-burn)*chains

# Get priors
finalModPriors <- get_prior(finalMod)
gpPrior <- finalModPriors |>
  filter(class == "lscale", source == "default") |>
  pull(prior) |>
  unique() |>
  str_subset(pattern = "inv_gamma")
lscalePriorAlpha <- as.numeric(str_extract(gpPrior, "(?<=inv_gamma\\()[^,]+"))
lscalePriorBeta <- as.numeric(str_extract(gpPrior, "(?<=, )[^)]+"))
```

To model the proportion of coral bleaching $Y_i$ for $i = 1, ..., 2835$, we use a **Bayesian Beta regression** with a **logit link function**:

$$
Y_i \sim \text{Beta}(\mu_i \phi, (1-\mu_i) \phi)
$$

where $\mu_i$ is the mean bleaching percentage, and $\phi$ is the precision parameter. Various models for the mean structure were examined and defined as:

<div style="padding-left:0px;padding-right:0px;padding-top:10px;padding-bottom:10px;overflow-x:auto;width:auto;">
$$
\begin{aligned}
\textbf{Model 1}: \text{logit}(\mu_i) &= \text{Date_Year}_i\beta_1 + \text{Lat}_i\beta_2 + \text{Lon}_i\beta_3 + \sum_{p} X_{ip}\beta_p \\
\textbf{Model 2}: \text{logit}(\mu_i) &= \text{Date_Year}_i\beta_1 + g(\text{Lat}, \text{Lon}) + \sum_{p} X_{ip}\beta_p \\
\textbf{Model 3}: \text{logit}(\mu_i) &= f(\text{Date_Year}) + \text{Lat}_i\beta_2 + \text{Lon}_i\beta_3 + \sum_{p} X_{ip}\beta_p \\
\textbf{Model 4}: \text{logit}(\mu_i) &= f(\text{Date_Year}) + g(\text{Lat}, \text{Lon}) + \sum_{p} X_{ip}\beta_p \\
\textbf{Model 5}: \text{logit}(\mu_i) &= f_{\text{City_Town_Name}}(\text{Date_Year}) + \text{Lat}_i\beta_2 + \text{Lon}_i\beta_3 + \sum_{p} X_{ip}\beta_p \\
\textbf{Model 6}: \text{logit}(\mu_i) &= f_{\text{City_Town_Name}}(\text{Date_Year}) + g(\text{Lat}, \text{Lon}) + \sum_{p} X_{ip}\beta_p \\
\end{aligned}
$$
</div>

where:

### Gaussian Process (GP) for Temporal Trends
<div style="padding-left:0px;padding-right:0px;padding-top:10px;padding-bottom:10px;overflow-x:auto;width:auto;">
$$
f_{\text{City_Town_Name}}(\text{Date_Year}) \sim \mathcal{GP} (0, (k_c(t_i, t_j))_{i,j = 1}^n) \\ 
$$
</div>

where the covariance function for each city $c$ is:

$$
k_c(t_i, t_j) = \sigma_c^2 \exp\left( -\frac{||t_i - t_j||^2}{2 \rho_c^2} \right)
$$

with:

* $t_i, t_j$ as observed `Date_Year` values.
* $c$ representing the city (`City_Town_Name`), where each city has a separate GP.
* $k_c(t_i, t_j)$ as the covariance function, using an exponentiated-quadratic (squared exponential) kernel.
* $\sigma_c^2$ representing a standard deviation parameter of the GP for city $c$.
* $\rho_c$ as the characteristic length-scale parameter.

### Tensor-Product Spline for Spatial Variation

$$
g(\text{Lat}, \text{Lon}) = \sum_{k_1} \sum_{k_2} \beta_{k_1 k_2} B_{k_1}(\text{Lat}) B_{k_2}(\text{Lon})
$$

where:

* $B_{k_1}(\text{Lat})$ and $B_{k_2}(\text{Lon})$ are basis functions for latitude and longitude.
* $\beta_{k_1 k_2}$ are the coefficients to be estimated.
* The smoothing penalty is controlled by a hyperparameter $\lambda$, which regularizes the estimated coefficients.

### Fixed Effects

$$
\sum_{p} X_{ip} \beta_p
$$

where:

* $X_{ip}$ is the value of the $p$-th covariate for observation $i$.
* $\beta_p$ is the corresponding regression coefficient.

### Prior Specification

* **Fixed Effects**: 
  + $\beta_p \sim \mathcal{N}(0,5)$ for all covariates $p$.
* **Gaussian Process (Temporal Trends)**:
  + $\sigma_c \sim \text{half-Cauchy}(0,2)$
  + $\rho_c \sim \text{InvGamma}(`r lscalePriorAlpha`, `r lscalePriorBeta`)$ (explicitly defined by `brms`)
* **Tensor-Product Spline**:
  + $\beta_{k_1 k_2} \sim \mathcal{N}(0,5)$
  + $\lambda \sim \text{half-Cauchy}(0,2)$ (if explicitly included in smoothing penalty)
* **Precision Parameter**: 
  + $\phi \sim \text{Gamma}(0.1, 0.1)$

This model accounts for both spatial and temporal dependencies, allowing for flexible trend estimation.

# Model Comparison

We tested the 6 models abive to evaluate different approaches for capturing spatiotemporal variation in coral bleaching. The candidate models included:

- **Linear models** with Date_Year as a fixed effect.
- **GP models**, both with and without city-specific trends.
- **Smoothed Spline models**, incorporating either Lat and Lon as fixed effects or a smooth spatial term.

After running convergence checks, the final model was selected using **Leave-One-Out Cross-Validation (LOO-CV)**, ensuring it provided the best balance between fit and complexity.

```{r Model Comparison}
# Load comparison data
load(file = "_data/fitCompDF.RData")

fitCompDF |>
  select(
    Model, 
    TempStr,
    SpaceStr, 
    elpd_diff_loo, 
    se_diff_loo,
    p_loo,
    looic, 
    se_looic
  ) |>
  arrange(desc(elpd_diff_loo)) |>
  gt() |>
  cols_label(
    TempStr = "Temporal Structure",
    SpaceStr = "Spatial Structure"
    # looic = "LOOIC",
    # se_looic = "SE LOOIC",
    # p_loo = "p-value",
    # elpd_diff_loo = "LOO ELPD Difference",
    # se_diff_loo = "SE LOO Difference"
  ) |>
  fmt_number(
    use_seps = FALSE
  ) |>
  tab_footnote(
    footnote = "Difference in Expected Log pointwise Predictive Density for a new dataset",
    locations = cells_column_labels(columns = elpd_diff_loo)
  ) |>
  tab_footnote(
    footnote = "Standard Error of component-wise elpd_diff_loo between two models",
    locations = cells_column_labels(columns = se_diff_loo)
  )  |>
  tab_footnote(
    footnote = "Effective number of parameters",
    locations = cells_column_labels(columns = p_loo)
  ) |>
  tab_footnote(
    footnote = "Leave-one-out Information Criteria",
    locations = cells_column_labels(columns = looic)
  ) |>
  tab_footnote(
    footnote = "Standard Error of looic",
    locations = cells_column_labels(columns = se_looic)
  ) |>
  tab_style(
    style = cell_text(whitespace = "nowrap"),
    locations = list(cells_body(), cells_column_labels())
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#373737"),
      cell_text(color = "#fff")
    ),
    locations = list(
      cells_column_labels()
    )
  ) |>
  tab_options(
    column_labels.padding = "10px",
    table.background.color = "#f2f2f2"
  ) |>
  #opt_vertical_padding(scale = 0.25) |>
  tab_caption(caption = md("Table 2: Model Comparison using LOO-CV")) |>
  as_raw_html()
```

## Selected Model

Model 6 emerged as the best-performing model in the comparison based on Leave-One-Out Information Criterion (LOOIC) and expected log predictive density (ELPD). It achieved the lowest LOOIC and the highest ELPD, indicating superior predictive accuracy while effectively balancing model complexity.

A key advantage of Model 6 was its flexible structure, incorporating:

* **City-Specific GPs** for temporal variation, capturing localized trends in bleaching over time.
* A **Tensor-Product Smoothed Spline** for spatial variation, allowing for smooth, nonlinear geographic effects.
* A broad set of environmental and physical predictors, including **Distance to Shore, Exposure, Turbidity, Cyclone Frequency, Depth, Windspeed, ClimSST, SSTA, TSA, and TSA_DHW**, hypothesized to drive bleaching dynamics.

Compared to alternative models, Model 6 provided the best trade-off between fit and generalizability, avoiding overfitting while preserving essential temporal and spatial dependencies. However, some covariates exhibited credible intervals overlapping zero, suggesting they might not contribute meaningfully. To enhance interpretability and model efficiency, we performed an iterative variable selection process, systematically removing weak predictors and reassessing model performance.

## Model Refinement and Variable Selection

To improve model parsimony and predictive performance, an iterative refinement process was conducted to remove covariates that did not contribute significantly to the model. The refinement process followed these steps:

1.  Identify Non-Significant Covariates

    * Variables whose 95% credible intervals contained zero were considered weak contributors.

2.  Iterative Variable Removal & Refitting

    * The least significant covariate was removed from the model. 
    * The model was then refit without that covariate to assess its impact. 

3.  Evaluate Model Fit via Bayes Factor & MAE

    * **Bayes Factor (BF) Comparison**: The refined model was compared to the previous iteration using `bayes_factor()`. If BF > 10, the new model was preferred. 
    * **LOOIC**: The reliability of how the refined model generalizes to new data was estimated. If LOOIC was lower, the new model was retained. 
    * **Mean Absolute Error** (MAE): The predictive performance was evaluated using the PPD from refined model compared to observed Percent Bleaching to check model improvement/degradation. If MAE improved or remained stable, the new model was retained.

4.  Repeat Until No Further Improvement

    * This process continued until all remaining covariates contributed meaningfully, ensuring the final model was both interpretable and robust.

Through this process, unnecessary covariates were systematically removed, leading to a final optimized model that retained only the most relevant predictors while maintaining strong predictive accuracy.

```{r Refinement}
load(file = "_data/refinementDF.RData")

refinementDF |>
  select(
    PriorModel, 
    RefinedModel,
    CovariateRemoved,
    BF,
    LOOIC,
    RefinedMAE
  ) |>
  gt() |>
  cols_label(
    PriorModel = "Prior Model",
    RefinedModel = "Refined Model",
    CovariateRemoved = "Covariate Removed",
    RefinedMAE = "MAE"
  ) |>
  sub_missing() |>
  fmt_number(
    columns = BF,
    decimals = 1,
    use_seps = FALSE
  ) |>
  fmt_number(
    columns = LOOIC,
    decimals = 1,
    use_seps = FALSE
  ) |>
  fmt_number(
    columns = RefinedMAE,
    decimals = 4
  ) |>
  tab_style(
    style = cell_text(whitespace = "nowrap"),
    locations = list(cells_body(), cells_column_labels())
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#373737"),
      cell_text(color = "#fff")
    ),
    locations = list(
      cells_column_labels()
    )
  ) |>
  tab_options(
    column_labels.padding = "10px",
    table.background.color = "#f2f2f2"
  ) |>
  #opt_vertical_padding(scale = 0.25) |>
  tab_caption(caption = md("Table 3: Model Refinement using BF, LOOIC, and MAE")) |>
  as_raw_html()
```

## Final Model

Through iterative model comparison, Model 9 was selected as the best-performing model, showing improvements over Model 6 in terms of fit and interpretability. This selection process involved removing non-significant covariates one at a time while assessing model performance metrics. Importantly, the smoothing parameters remained unchanged throughout this refinement process, ensuring consistency in spatial and temporal trends. The final model captures essential environmental and climatic predictors, balancing complexity and generalizability.

Model 9 includes key predictors such as Distance to Shore, Turbidity, Cyclone Frequency, Windspeed, Sea Surface Temperature Anomalies (SSTA and TSA), and Degree Heating Weeks derived from TSA (TSA_DHW). These covariates were retained based on their statistical significance and their ecological relevance to coral bleaching dynamics. The refined model structure provides a robust framework for understanding and predicting bleaching patterns, facilitating targeted conservation efforts.

# Goodness of Fit

A key aspect of evaluating the selected model’s reliability is examining its ability to replicate observed data patterns. Posterior predictive checks provide a direct way to assess the extent to which simulations from the model align with the actual observed data.

## Posterior Predictive Checks

To evaluate the model’s fit, we conducted posterior predictive checks (PPCs), which compare the observed data to simulated draws from the posterior predictive distribution. The following visualizations assess whether the model-generated data resemble the observed coral bleaching percentages.

```{r PPC}
### PPC ----
# Makes draws for plots
set.seed(52) # for reproducibility
posteriorFinalMod <- posterior_predict(finalMod)

# Observed data
## Response
obsY <- bleachingData$PercentBleachingBounded

## Grouping
groupCity <- bleachingData$City_Town_Name

# Random draws
numDraws <- 400
set.seed(52) # for reproducibility
drawsInd <- sample(x = 1:sims, size = numDraws)

# Posterior sample
postYsamp <- posteriorFinalMod[drawsInd, ]

#### Plot guide 
fillPPC <- "#d1e1ec"
colorPPC <- "#b3cde0"
fill2PPC <- "#011f4b"
```

### Distribution Overlay

Figure 4 presents an overlay of the posterior predictive distribution (PPD) against the observed bleaching percentages. The solid black line represents the observed data ($y$), while the blue-shaded posterior simulations ($y_{rep}$) provide an indication of model uncertainty. The strong alignment between the observed and predicted densities suggests that the model successfully captures the overall distribution of coral bleaching percentages.

```{r PPC Distribution, fig.align='center', out.width="90%", fig.cap="Figure 4: Posterior Predictive Distribution vs Observed Percent Bleaching from 400 posterior draws"}
#### Density ----
set.seed(52) # for reproducibility
ppcDensPlot <- ppc_dens_overlay(
  y = obsY,
  #yrep = posteriorFinalMod
  yrep = postYsamp
) +
  scale_x_continuous(
    name = "Percent Bleaching",
    limits = c(0,1),
    breaks = seq(0, 1, 0.1)
  ) +
  scale_y_continuous(
    name = "Density",
    expand = expansion(mult = c(0, 0.01))
  ) +
  labs(
    #title = "Posterior Predictive Distribution vs Observed Percent Bleaching",
    #subtitle = paste(sims, "Simulations")
  )
ppcDensPlot
```

### Distributional Statistics

The set of plots in Figure 5 evaluates how well the model reproduces key summary statistics of the observed data, including:

* Mean
* Standard deviation (SD)
* 2.5% Lower credible bound (LCB)
* Median
* 97.5% Upper credible bound (UCB)

Each histogram represents the distribution of these statistics across 8000 posterior simulations, with the vertical black line indicating the observed statistic. The Bayesian p-values assess whether the observed value is typical less than the posterior predictive distribution values. Values close to 0.5 suggest a good fit, while values near 0 or 1 may indicate potential discrepancies.

Overall, these diagnostics confirm that the final model provides a reasonable approximation of the observed data, supporting its validity for inference and prediction.

```{r PPC Stats, fig.align='center', out.width = "90%", fig.cap = "Figure 5: Posterior Predictive Checks for Distributional Statistics with Bayesian Predictive p-values from 8000 Simulations"}
#### Stats ----
# Make stat functions
meanFunc <- function(y){mean(y)}
sdFunc <- function(y){sd(y)}
medianFunc <- function(y){median(y)}
lcbFunc <- function(y){quantile(y, 0.025)}
ucbFunc <- function(y){quantile(y, 0.975)}

##### Mean ----
set.seed(52) # for reproducibility
meanY <- meanFunc(obsY)

ppcMeanStat <- ppc_stat_data(
  y = obsY,
  yrep = posteriorFinalMod,
  group = NULL,
  stat = c("meanFunc")
) |>
  mutate(
    meanProbLow = value < meanY,
    meanProbHigh = value > meanY
  )

ppcMeanPlotGG <- ggplot() +
  geom_histogram(
    data = ppcMeanStat |> filter(variable != "y"),
    aes(x = value, color = "Posterior"),
    fill = fillPPC
  ) +
  geom_vline(
    data = ppcMeanStat |> filter(variable == "y"),
    aes(xintercept = value, color = "Observed"),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = "Percent Bleaching"
  ) +
  scale_y_continuous(
    name = "Number of Posterior Draws",
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_color_manual(
    name = "Data",
    values = c(
      "Posterior" = colorPPC,
      "Observed" = "black"
    ),
    breaks = c("Posterior", "Observed")
  ) +
  labs(title = "Mean",
       subtitle = paste("p-value =", round(mean(ppcMeanStat$meanProbLow[-1]), 4))
  ) +
  theme_bw() +
  guides(color = guide_legend(byrow = TRUE)) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.key.spacing.y = unit(5, "points"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

##### SD ----
set.seed(52) # for reproducibility
sdY <- sdFunc(obsY)

ppcSDStat <- ppc_stat_data(
  y = obsY,
  yrep = posteriorFinalMod,
  group = NULL,
  stat = c("sdFunc")
) |>
  mutate(
    sdProbLow = value < sdY,
    sdProbHigh = value > sdY
  )

ppcSDPlotGG <- ggplot() +
  geom_histogram(
    data = ppcSDStat |> filter(variable != "y"),
    aes(x = value, color = "Posterior"),
    fill = fillPPC
  ) +
  geom_vline(
    data = ppcSDStat |> filter(variable == "y"),
    aes(xintercept = value, color = "Observed"),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = "Percent Bleaching"
  ) +
  scale_y_continuous(
    name = "Number of Posterior Draws",
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_color_manual(
    name = "Data",
    values = c(
      "Posterior" = colorPPC,
      "Observed" = "black"
    ),
    breaks = c("Posterior", "Observed")
  ) +
  labs(title = "SD",
       subtitle = paste("p-value =", round(mean(ppcSDStat$sdProbLow[-1]), 4))
  ) +
  theme_bw() +
  guides(color = guide_legend(byrow = TRUE)) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.key.spacing.y = unit(5, "points"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

##### Median ----
set.seed(52) # for reproducibility
medianY <- medianFunc(obsY)

ppcMedianStat <- ppc_stat_data(
  y = obsY,
  yrep = posteriorFinalMod,
  group = NULL,
  stat = c("medianFunc")
) |>
  mutate(
    medianProbLow = value < medianY,
    medianProbHigh = value > medianY
  )

ppcMedianPlotGG <- ggplot() +
  geom_histogram(
    data = ppcMedianStat |> filter(variable != "y"),
    aes(x = value, color = "Posterior"),
    fill = fillPPC
  ) +
  geom_vline(
    data = ppcMedianStat |> filter(variable == "y"),
    aes(xintercept = value, color = "Observed"),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = "Percent Bleaching"
  ) +
  scale_y_continuous(
    name = "Number of Posterior Draws",
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_color_manual(
    name = "Data",
    values = c(
      "Posterior" = colorPPC,
      "Observed" = "black"
    ),
    breaks = c("Posterior", "Observed")
  ) +
  labs(title = "Median",
       subtitle = paste("p-value =", round(mean(ppcMedianStat$medianProbLow[-1]), 4))
  ) +
  theme_bw() +
  guides(color = guide_legend(byrow = TRUE)) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.key.spacing.y = unit(5, "points"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

##### LCB ----
set.seed(52) # for reproducibility
lcbY <- lcbFunc(obsY)

ppcLCBStat <- ppc_stat_data(
  y = obsY,
  yrep = posteriorFinalMod,
  group = NULL,
  stat = c("lcbFunc")
) |>
  mutate(
    lcbProbLow = value < lcbY,
    lcbProbHigh = value > lcbY
  )

ppcLCBPlotGG <- ggplot() +
  geom_histogram(
    data = ppcLCBStat |> filter(variable != "y"),
    aes(x = value, color = "Posterior"),
    fill = fillPPC
  ) +
  geom_vline(
    data = ppcLCBStat |> filter(variable == "y"),
    aes(xintercept = value, color = "Observed"),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = "Percent Bleaching"
  ) +
  scale_y_continuous(
    name = "Number of Posterior Draws",
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_color_manual(
    name = "Data",
    values = c(
      "Posterior" = colorPPC,
      "Observed" = "black"
    ),
    breaks = c("Posterior", "Observed")
  ) +
  labs(title = "LCB",
       subtitle = paste("p-value =", round(mean(ppcLCBStat$lcbProbLow[-1]), 4))
  ) +
  theme_bw() +
  guides(color = guide_legend(byrow = TRUE)) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.key.spacing.y = unit(5, "points"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

##### UCB ----
set.seed(52) # for reproducibility
ucbY <- ucbFunc(obsY)

ppcUCBStat <- ppc_stat_data(
  y = obsY,
  yrep = posteriorFinalMod,
  group = NULL,
  stat = c("ucbFunc")
) |>
  mutate(
    ucbProbLow = value < ucbY,
    ucbProbHigh = value > ucbY
  )

ppcUCBPlotGG <- ggplot() +
  geom_histogram(
    data = ppcUCBStat |> filter(variable != "y"),
    aes(x = value, color = "Posterior"),
    fill = fillPPC
  ) +
  geom_vline(
    data = ppcUCBStat |> filter(variable == "y"),
    aes(xintercept = value, color = "Observed"),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = "Percent Bleaching"
  ) +
  scale_y_continuous(
    name = "Number of Posterior Draws",
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_color_manual(
    name = "Data",
    values = c(
      "Posterior" = colorPPC,
      "Observed" = "black"
    ),
    breaks = c("Posterior", "Observed")
  ) +
  labs(title = "UCB",
       subtitle = paste("p-value =", round(mean(ppcUCBStat$ucbProbLow[-1]), 4))
  ) +
  theme_bw() +
  guides(color = guide_legend(byrow = TRUE)) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.key.spacing.y = unit(5, "points"),
    legend.position = "bottom",
    legend.direction = "horizontal"
  )

#### Combined plot ----
ppcCombPlot <- 
  (ppcMeanPlotGG + ppcSDPlotGG) /
  (ppcLCBPlotGG + ppcMedianPlotGG + ppcUCBPlotGG) +
  plot_layout(
    guides = "collect",
    axes = "collect_x"
  ) +
  plot_annotation(
    #title = "Posterior Predictive Checks for Distributional Statistics",
    #subtitle = paste("Bayesian predictive p-values for", sims, "Simulations"),
    theme = theme(
      legend.position = "bottom",
      legend.direction = "horizontal"
    )
  )
ppcCombPlot
```

# Model Results

The final model's results offer insights into the factors influencing coral bleaching percentages.

## Variable Importance

1.  Parameter Estimates  
The table below presents the estimated coefficients, standard errors, and 95% credible intervals for each covariate:

```{r Model Results Table}
fixedEff <- fixef(finalMod) |>
  data.frame() |>
  mutate(
    p_val = dnorm(Estimate/Est.Error)
  ) |>
  mutate(
    across(everything(), function(x){round(x, 4)})
  ) |>
  mutate(
    Sig = ifelse(p_val < 0.01, "***",
                 ifelse(p_val < 0.05, "**",
                        ifelse(p_val < 0.1, "*", "")))
  )

# Render Table
fixedEff |>
  rownames_to_column(var = "Parameter") |>
  select(
    Parameter,
    Estimate,
    Est.Error,
    Q2.5,
    Q97.5
  ) |>
  filter(str_detect(Parameter, "t2", negate = TRUE)) |>
  gt() |>
  cols_label(
    Estimate = "{{:beta:}}",
    Est.Error = "SD({{:beta:}})"
  ) |>
  cols_merge(
    columns = c(Q2.5, Q97.5), pattern = "({1}, {2})"
  ) |>
  cols_label(
    Q2.5 = "95% CI"
  ) |>
  cols_align(
    columns = Q2.5,
    align = "center"
  ) |>
  tab_footnote(
    footnote = "Parameter estimate",
    locations = cells_column_labels(columns = Estimate)
  ) |>
  tab_footnote(
    footnote = "Standard Deviation of parameter estimate",
    locations = cells_column_labels(columns = Est.Error)
  ) |>
  tab_footnote(
    footnote = "95% Credible Interval of parameter estimate",
    locations = cells_column_labels(columns = Q2.5)
  ) |>
  tab_style(
    style = cell_text(whitespace = "nowrap"),
    locations = list(cells_body(), cells_column_labels())
  ) |>
  tab_style(
    style = list(
      cell_text(
        align = "center"
      )
    ),
    locations = list(
      cells_column_labels(
        columns = c(Estimate, Est.Error)
      )
    )
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#373737"),
      cell_text(color = "#fff")
    ),
    locations = list(
      cells_column_labels()
    )
  ) |>
  tab_options(
    column_labels.padding = "10px",
    table.background.color = "#f2f2f2"
  ) |>
  #opt_vertical_padding(scale = 0.25) |>
  tab_caption(caption = md("Table 4: Parameter Estimates")) |> 
  as_raw_html() |> I()
```

  \noindent Interpretation: 

    * Distance to Shore: A positive coefficient suggests that as the distance from shore   increases, the percentage of coral bleaching also increases. 
    * Turbidity: The negative coefficient indicates that higher turbidity (murkier water) is associated with a decrease in coral bleaching percentages.  
    * Windspeed: Higher wind speeds are linked to reduced coral bleaching.  
    * SSTA (Sea Surface Temperature Anomaly): A slight increase in SSTA correlates with a   decrease in bleaching, though the effect size is small. 
    * TSA (Tropical Sea Surface Temperature Anomaly): Higher TSA values are associated with increased coral bleaching.  
    * TSA_DHW (Degree Heating Weeks derived from TSA): Elevated TSA_DHW corresponds to higher bleaching percentages.  

2.  Smoothing Spline and Gaussian Process Hyperparameters: 
The model incorporates spatial and temporal random effects to account for variability across different locations and times:  

    * Spatial Effects (t2(Lat, Lon)): The estimated standard deviations for the spatial components suggest variability in bleaching across different latitudes and longitudes.
    * Temporal Effects (gp(Date_Year) by City_Town_Name): The Gaussian Process terms capture temporal trends within each city or town, indicating that bleaching patterns change over time differently across locations.  

3.  Precision Parameter (phi):
The estimated precision parameter (phi) is 8.8059, with a standard error of 0.2647. A higher phi value indicates that the data points are closely clustered around the mean, suggesting less variability in bleaching percentages after accounting for the covariates.




